The Astropy Project

### 2017-12-08 main.tex

The development of the \astropypkg core package began as a largely
community-driven effort to standardize core functionality for astronomical
software in \python.
In this way, its genesis differs from but builds upon many substantial and
former astronomical software development efforts that were commissioned or
initiated through large institutional support, for example IRAF \citep[developed
at NOAO;][]{IRAF}, MIDAS \citep[developed at ESO;][]{MIDAS}, or Starlink
\citep[originally developed by a consortium of UK institutions and now
maintained by the East Asian Observatory;][]{starlink1982,starlink2013}.
More recently, community-driven efforts have seen significant success in the astronomical sciences \citep{yt}.

\python\footnote{\url{https://www.python.org/}} is an increasingly popular, general-purpose
programming language that is available under a permissive open source software license free of
charge for all major operating systems. The programming language has become especially popular
in the quantitative sciences, where researchers must simultaneously produce research, perform
data analysis, and develop software. A large part of this success owes itself to the vibrant
community of developers and a continuously-growing ecosystem of tools, web services, and stable
well-developed packages that enable easier collaboration on software development, easier
writing and sharing of software documentation, and continuous testing and validation of
software. While dedicated libraries provide support for array representation and
arithmetic \citep[\package{numpy};][]{numpy}, a wide variety of functions for scientific
computing \citep[\package{scipy};][]{scipy}, and publication-quality plotting
\citep[\package{matplotlib};][]{matplotlib}, tens of thousands of other high-quality and easy-to-use
packages are available, which can help with tasks that are not astronomy specific but
might be performed in the course of astronomical research; e.g., interfacing with
databases or statistical inferences. More recently, the development of package managers such as
Anaconda\footnote{\url{https://anaconda.org/}} has significantly streamlined the installation
process for many libraries, lowering many barriers to entry.

@@

The \astropy project aims to develop and provide high-quality code and
documentation according to the best practices in software development.
The project makes use of different tools and web services to reach those
goals without central institutional oversight.
The first public release of the \astropypkg package is described in
\cite{astropy}. Since then, the \astropypkg package has been
used in hundreds of projects and the scope of the package has grown
considerably. At the same time, the scientific community
contributing to the project has grown tremendously and an ecosystem
of packages supporting or affiliated with the \astropypkg core has
developed.
In this paper, we describe the current status of the \astropy community and the
\astropypkg core package and discuss goals for future development.

@@

For resolving disagreements about the \astropypkg core package or other \astropy-managed code, the coordination committee primarily acts to work toward consensus, or when consensus is difficult to achieve, generally acts as a ``tie-breaker.''
The committee also oversees affiliated package applications to ensure they are
in keeping with \astropy's vision and
philosophy,\footnote{\url{http://docs.astropy.org/en/stable/development/vision.html}}
as well as the associated procedures.
Additionally, the committee oversees the assignment of roles (primarily driven by already-existing contributions), and increasingly has acted as the ``face'' of the Project, providing contact with organizations like NumFOCUS (the body that holds any potential funding in trust for \astropy) or the American Astronomical Society (AAS).

@@

Central to the success of \astropy is an open environment where anybody can
contribute to the project.
This model leads to an ``organic'' growth, where different features are
implemented by different people with different programming styles and
interfaces.
Thus, \astropy has a mechanism to more formally propose significant changes to
the core package (e.g., re-writing the coordinates subpackage; \citealt{ape5}),
to plan out major new features (e.g., a new file format; \citealt{ape6}), or
institute new organization-wide policies (e.g., adopting a code of conduct;
\citealt{ape8}).
This mechanism is called ``Astropy Proposal for Enhancement'' (APE) and is
modeled after the ``Python Enhancement Proposals'' (PEP) that guide the
development of the \python programming language.
In an APE, one or more authors describe in detail the proposed changes or
additions, including a rationale for the changes, how these changes will be
implemented, and in the case of code, what the interface will be \citep{ape1}.
The APEs are discussed and refined by the community before much work is invested
into a detailed implementation; anyone is welcome to contribute to these
discussions during the open consideration period. APEs are proposed via pull
requests on a dedicated GitHub repository\footnote{\url{https://github.com/astropy/astropy-APEs}};
anyone can therefore read the proposed APEs and leave in-line comments.
When a community consensus emerges, the APEs are accepted and become
the basis for future work.
In cases where consensus cannot be reached, the
\astropy coordination committee may decide to close the discussion and
make an executive decision based on the community input on the APE in question.

@@

The \astropypkg package has a regular release schedule consisting of new significant
releases every 6 months, with bugfix releases as needed \citep{ape2}.
The major releases contain new features or any significant changes, whereas
the bugfix releases only contain fixes to code or documentation but no new
features.
Some versions are additionally designated as ``Long-term support'' (LTS)
releases, which continue to receive bug fixes for 2 years following the release
with no changes to the API\@.
The LTS versions are ideal for pipelines and other applications, where API
stability is essential.
The latest LTS release (version 2.0) is also the last one that supports \python
2; it will receive bug fixes until the end of 2019 \citep{ape10}.

@@

The released versions of the \astropypkg core package are available from several
of the \python distributions for scientific computing (e.g.,
\href{http://anaconda.org}{Anaconda}) and from the \python Package Index
(PyPI).\footnote{See the installation documentation for more information:
\url{http://docs.astropy.org/en/stable/install.html}}
Effort has been made to make \astropypkg available and easily installable across
all platforms; the package is constantly tested on different platforms as part
of a suite of continuous integration (CI) tests.

@@

The \astropy project, as of the version 2.0 release, does not receive any direct
financial support for the development of \astropypkg.
Development of the software, all supporting materials, and community support are
provided by individuals who work on the \astropy project in their own personal
time, by individuals or groups contributing to \astropy as part of a research
project, or contributions from institutions that allocate people to work on
\astropy.
A list of organizations that have contributed to \astropy in this manner
can be found in the Acknowledgments.

@@

\section{Astropy Core Package Version 2.0}
\label{sec:core}
% Adrian
The \astropy project aims to provide \python-based packages for all tasks that
are commonly needed in a large subset of the astronomical community.
At the foundation is the \astropypkg core package, which provides general
functionality (e.g., coordinate transformations, reading and writing
astronomical files, and units) or base classes for other
packages to utilize for a common interface (e.g., \texttt{NDData}).
In this section, we highlight new features introduced or substantially improved
since version 0.2 (previously described in \citealt{astropy}).  The \astropypkg
package
provides a full log of changes\footnote{\url{https://github.com/astropy/astropy/blob/stable/CHANGES.rst}}
over the course of the entire project and more details about individual
subpackages are available in the documentation.\footnote{\url{http://docs.astropy.org/en/stable/}}
Beyond what is mentioned below, most subpackages have seen increased performance
since the release of the version 0.2 package.

@@

As of \astropypkg version 2.0, units and quantities, prevalent in most of its
subpackages, have become a key concept for using the package as a whole.
Units are intimately entwined in the definition of astronomical coordinates;
thus, nearly all functionality in the \texttt{astropy.coordinates} subpackage
(see \sectionname~\ref{sec:coordinates}) depends on them.
For most other subpackages, quantities are at least accepted and often expected
by default.

@@

@@

\subsubsection{Defining functions that require quantities}
        When writing code or
        functions that expect \texttt{Quantity} objects, we often want to
        enforce that the input units have the correct physical type.
        For example, we may want to require only length-type \texttt{Quantity}
        objects.
        \texttt{astropy.units} provides a tool called \texttt{quantity\_input()}
        that can perform this verification automatically to avoid repetitive
        code.

@@

The classes that handle coordinate representations (the \texttt{Representation}
classes) act like three-dimensional vectors and thus support vector arithmetic.
The classes that represent reference systems and frames (the \texttt{Frame}
classes) internally use \texttt{Representation} objects to store the coordinate
data---that is, the \texttt{Frame} classes accept coordinate data, either as a
specified \texttt{Representation} object, or using short-hand keyword arguments
to specify the components of the coordinates.
These preferred representation and short-hand component names differ between
various astronomical reference systems.
For example, in the ICRS frame, the spherical angles are right ascension
(\texttt{ra}) and declination (\texttt{dec}), whereas in the Galactic frame, the
spherical angles are Galactic longitude (\texttt{l}) and latitude (\texttt{b}).
Each \texttt{Frame} class defines its own component names and preferred \texttt{Representation} class.
The frame-specific component names map to corresponding components on the
underlying \texttt{Representation} object that internally stores the coordinate
data.
For most frames, the preferred representation is spherical, although this is
determined primarily by the common use in the astronomical community.

@@

\begin{figure}
\includegraphics[width=\textwidth]{coordinates_graph.pdf}
\caption{%
    The full graph of possible reference frame transformations implemented in
    \texttt{astropy.coordinates}.
    Arrows indicate transformations from one frame to another.
    Arrows that point back to the same frame indicate self-transformations that
    involve a change of reference frame parameters (e.g., equinox).
    \label{fig:frame-transform-graph}
}
\end{figure}

@@

\subsubsection{Proper motion and velocity transformations}
    In addition to positional coordinate data, the \texttt{Frame} classes now
    also support velocity data.
    As the default representation for most frames is spherical, most of the
    \texttt{Frame} classes expect proper motion and radial velocity components
    to specify the velocity information.
    The names of the proper motion components all start with \texttt{pm} and
    adopt the same longitude and latitude names as the positional components.
    Transforming coordinates with velocity data is also supported, but in some
    cases the transformed velocity components have limited accuracy because the
    transformations are done numerically instead of analytically.
    The low-level interface for specifying and transforming velocity data (see
    the next point) is currently experimental.
    As such, in version 2.0, only the \texttt{Frame} classes (and not the
    \texttt{SkyCoord} class) support handling velocities.

\subsubsection{Solar System Ephemerides}
    Also new is support for computing ephemerides of major solar system bodies
    and outputting the resulting positions as coordinate objects.
    These ephemerides can be computed either using analytic approximations from
    ERFA or from downloaded JPL ephemerides (the latter requires the
    \package{jplephem}\footnote{\url{https://github.com/brandon-rhodes/python-jplephem}}
    optional dependency and an internet connection).

@@

\tablename~\ref{tab:coordinate_benchmarks} gives an example of the relative
accuracy of the conversion from FK4 to Galactic coordinates for all pairs of
tools. This shows, for example, that \astropy, Kapteyn, and PyTPM agree
perfectly (to the precision shown), while PALpy, pySLALIB, and PyAST also
agree perfectly amongst themselves. However, there is an offset of around
0.2\arcsec\space with the former three packages. Finally, PyEphem disagrees
with other packages by 0.4--0.8\arcsec. These values are only meant to be
illustrative and will change over time as the benchmarks are refined and the
packages updated.
%\inlinecomment{Perry G}{Wouldn't the table be better represented as a matrices? Or perhaps as a matrix as a crude image with colors or intensities representing the size of the quantity. Is the difference between median and mean significant enough to warrant showing both? A link to the actual numbers may satisfy readers that want the exact values, but I suspect the majority don't need to see that precision.}
\begin{table}
\begin{center}
\begin{tabular}{llcccc}
\hline
\hline
Package 1 & Package 2 & Median & Mean & Maximum & Std. Dev. \\
          &           & arcsec & arcsec & arcsec & arcsec \\
\hline
astropy  &  kapteyn  & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata  &  pytpm    & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata  &  palpy    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyast    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyslalib & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyephem  & 0.459  & 0.458 & 0.860 & 0.231 \\
kapteyn  &  pytpm    & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata  &  palpy    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyast    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyslalib & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyephem  & 0.459  & 0.458 & 0.860 & 0.231 \\
pytpm    &  palpy    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyast    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyslalib & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyephem  & 0.459  & 0.458 & 0.860 & 0.231 \\
palpy    &  pyast    & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata  &  pyslalib & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata  &  pyephem  & 0.563  & 0.570 & 1.012 & 0.253 \\
pyast    &  pyslalib & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata  &  pyephem  & 0.563  & 0.570 & 1.012 & 0.253 \\
pyslalib &  pyephem  & 0.563  & 0.570 & 1.012 & 0.253 \\
\hline
\end{tabular}
\end{center}
\caption{Comparison of the accuracy of the FK4 to Galactic transformation between different packages.\label{tab:coordinate_benchmarks}}
\end{table}

\begin{figure}
  \includegraphics[width=\textwidth]{coordinates-benchmark.pdf}
  \caption{
    Comparison matrix of the accuracy values shown on
    table~\ref{tab:coordinate_benchmarks} of the FK4 to Galactic transformation
    between the different packages. Each comparison block shows four squares
    representing the median, mean, maximum and standard deviation going from
    left to right and bottom to top.
    \label{fig:coordinate_benchmarks}
  }
\end{figure}

@@

The \texttt{NDDataBase} class provides the abstract interface for gridded data
with attributes for accessing metadata, the world coordinate system (WCS),
uncertainty arrays matched to the shape of the data, and other traits.
Building on this interface, the \texttt{NDData} class provides a minimal
working implementation for storing \package{numpy} arrays. These classes serve
as useful base classes for package authors wishing to develop their own classes
for specific use cases and as containers for exchanging gridded data.

@@

The \package{astropy.table} subpackage provides functionality for representing
and manipulating heterogeneous data.
The package was described in detail in \cite{astropy}.
Thus, in what follows, we only summarize key new features or updates to
\package{astropy.table} since the previous \astropy paper.

@@

A table can contain data that naturally form groups; for example, it may
contain multiple observations of a few sources at different points in time
and in different bands. Then, we may want to split the table into groups based
on the combination of source observed and the band, after which we combine the
results for each combination of source and band in some way (e.g., finding
the mean or standard deviation of the fluxes or magnitudes over time) or filter
the groups based on user-defined criteria. These kinds of grouping and
aggregation operations are now fully supported by \texttt{Table} objects.

@@

\subsubsection{Using Astropy objects in tables}

The \texttt{Table} object now allows \texttt{Quantity}, celestial
coordinate (\texttt{SkyCoord}), and date/time (\texttt{Time}) objects to
be used as columns. It also provides a general way for other user-defined
objects to be used as columns. This makes it possible, for instance, to easily
represent catalogs of sources or time series in \astropy, while having both the
benefits of the \texttt{Table} object (e.g., accessing specific rows/columns
or groups of them and combining tables) and of, for example,
the \texttt{SkyCoord} or the \texttt{Time} classes (e.g., converting the
coordinates to a different frame or accessing the date/time in the desired time scale).

@@

The \package{astropy.io} subpackage provides support for reading and writing
data to a variety of ASCII and binary file formats, such as a wide range of
ASCII data table formats, FITS, and VOTable.
It also provides a unified interface for reading and writing data with these
different formats using the \package{astropy.table} subpackage.
For many common cases, this simplifies the process of file input and output (I/O) and
reduces the need to master the separate details of all the I/O packages within
\astropypkg.

\subsubsection{ASCII}

@@

The \package{astropy.io.ascii} subpackage now includes a significantly faster
Cython/C engine for reading and writing ASCII files. This is available for most
of the common formats.  On average, the new engine is about 4 to 5 times faster
than the corresponding pure-\python implementation and is often comparable to
the speed of the \package{pandas} \citep{pandas} ASCII file interface.  The
fast reader has parallel processing option that allows harnessing multiple
cores for input parsing to achieve even greater speed gains.  By default,
\texttt{read()} and \texttt{write()} will attempt to use the fast Cython/C engine
when dealing with compatible formats. Certain features of the full read
/ write interface are unavailable in the fast version, in which case the
pure-\python version will automatically be used.

The \package{astropy.io.ascii} subpackage now provides the capability
to read a table within an HTML file or web URL into an \astropypkg
\texttt{Table} object. A \texttt{Table} object can now also
be written out as an HTML table.

@@

\package{matplotlib} allows a custom normalization and stretch to be used when
displaying data by passing in a normalization object.
The \package{astropy.visualization} package also provides a normalization class
that wraps the interval and stretches objects into a normalization object that
\package{matplotlib} understands.

>>>>

The development of the \astropypkg core package began as a largely
community-driven effort to standardize core functionality for astronomical
software in \python.
In this way, its genesis differs from but builds upon many substantial and
% * <bsipocz@gmail.com> 2017-11-13T13:13:08.884Z:
%
% how do astroy builds upon these? or it's more like builds upon the developer experience?
%
%
% ^.
former astronomical software development efforts that were commissioned or
initiated through large institutional support, for example IRAF \citep[developed
at NOAO;][]{IRAF}, MIDAS \citep[developed at ESO;][]{MIDAS}, or Starlink
\citep[originally developed by a consortium of UK institutions and now
maintained by the East Asian Observatory;][]{starlink1982,starlink2013}.
More recently, community-driven efforts have seen significant success in the astronomical sciences \citep{yt}.

\python\footnote{\url{https://www.python.org/}} is an increasingly popular, general-purpose
programming language that is available under a permissive open source software license free of
charge for all major operating systems. The programming language has become especially popular
in the quantitative sciences, where researchers must simultaneously produce research, perform
data analysis, and develop software. A large part of this success owes itself to the vibrant
community of developers and a continuously-growing ecosystem of tools, web services, and stable
well-developed packages that enable easier collaboration on software development, easier
writing and sharing of software documentation, and continuous testing and validation of
software. While the programming language provides support for array representation and
arithmetic \citep[\package{numpy};][]{numpy}, a wide variety of functions for scientific
computing \citep[\package{scipy};][]{scipy}, and publication-quality plotting
\citep[\package{matplotlib};][]{matplotlib}, tens of thousands of other high-quality and easy-to-use
packages are available, which can help with tasks that are not astronomy specific but
might be performed in the course of astronomical research; e.g., interfacing with
databases or statistical inferences. More recently, the development of package managers such as
Anaconda\footnote{\url{https://anaconda.org/}} have streamlined the installation process for
most packages, significantly reducing the entry barrier for using many such libraries.

@@

The \astropy project aims to develop and provide high-quality code and
documentation according to the best practices in software development.
The project makes use of these tools to do so without central institutional
oversight.
The first public release of the \astropypkg package is described in
\cite{astropy}. Since then, the \astropypkg package has been
used in hundreds of projects and the scope of the package has grown
considerably. At the same time, the scientific community
contributing to the project has grown tremendously and an ecosystem
of packages supporting or affiliated with the \astropypkg core has
developed.
In this paper, we describe the current status of the \astropy community and the
\astropypkg core package and discuss goals for future development.

@@

For resolving disagreements about the \astropypkg core package or other \astropy-managed code, the coordination committee primarily acts to work toward consensus, or when consensus is difficult to achieve, generally acts as a ``tie-breaker.''
The committee also oversees affiliated package applications to ensure they are in keeping with \astropy's vision and philosophy, as well as the associated procedures.
Additionally, the committee oversees the assignment of roles (primarily driven by already-existing contributions), and increasingly has acted as the ``face'' of the Project, providing contact with organizations like NumFOCUS (the body that holds any potential funding in trust for \astropy) or the American Astronomical Society (AAS).

@@

Central to the success of \astropy is an open environment where anybody can
contribute to the project.
This model leads to an ``organic'' growth, where different features are
implemented by different people with different programming styles and
interfaces.
Thus, \astropy has a mechanism to more formally propose significant changes to
the core package (e.g., re-writing the coordinates subpackage; \citealt{ape5}),
to plan out major new features (e.g., a new file format; \citealt{ape6}), or
institute new organization-wide policies (e.g., adopting a code of conduct;
\citealt{ape8}).
This mechanism is called ``Astropy Proposal for Enhancement'' (APE) and are
modeled after the ``Python Enhancement Proposals'' (PEP) that guide the
development of the \python programming language.
In an APE, one or more authors describe in detail the proposed changes or
additions, including a rationale for the changes, how these changes will be
implemented, and in the case of code, what the interface will be \citep{ape1}.
The APEs are discussed and refined by the community before much work is invested
into a detailed implementation; anyone is welcome to contribute to these
discussions during the open consideration period. APEs are proposed via pull
requests on a dedicated GitHub repository\footnote{\url{https://github.com/astropy/astropy-APEs}};
anyone can therefore read the proposed APEs and leave in-line comments.
When a community consensus emerge, the APEs are accepted and become
the basis for future work.
In cases where consensus cannot be reached, the
\astropy coordination committee may decide to close the discussion and
make an executive decision based on the community input on the APE in question.

@@

The \astropypkg package has a regular release schedule consisting of new significant
releases every 6 months, with bugfix releases as needed \citep{ape2}.
The major releases contain new features or any significant changes, whereas
the bugfix releases only contain fixes to code or documentation but no new
features.
Some versions are additionally designated as ``Long-term support'' (LTS)
releases, which continue to receive bug fixes for 2 years following the release
with no changes to the API\@.
The LTS versions are ideal for pipelines and other applications, where API
stability is essential.
The latest LTS release (v2.0) is also the last one that supports \python 2; It
will receive bug fixes until the end of 2019 \citep{ape10}.

@@

The released versions of the \astropypkg core package are available from several
of the \python distributions for scientific computing (e.g.,
\href{http://anaconda.org}{Anaconda}) and from the \python Package Index
(PyPI).\footnote{See the installation documentation for more information:
\url{http://docs.astropy.org/en/stable/install.html}}
Effort has been made to make \astropypkg available and easily installable across
all platforms; the package is constantly tested on different platforms as part
of our suite of continuous integration tests.

@@

The \astropy project, as of the v2.0 release, does not receive any direct financial support for
the development of \astropypkg.  Development of the software, all supporting materials, and community support are provided
by individuals who work on the \astropy project in their own personal
time, by individuals or groups contributing to \astropy as
part of a research project, or contributions from institutions that allocate
people to work on \astropy.
A list of organizations that have contributed to \astropy in this manner
can be found in the Acknowledgements.

@@

\section{Astropy Core Package v2.0}
\label{sec:core}
% Adrian
The \astropy project aims to provide \python-based packages for all tasks that
are commonly needed in a large subset of the astronomical community.
At the foundation is the \astropypkg core package, which provides general
functionality (e.g., coordinate transformations, reading and writing
astronomical files, and units) or base classes for other
packages to utilize for a common interface (e.g., \texttt{NDData}).
In this section, we highlight new features introduced or substantially improved
since version v0.2 (previously described in \citealt{astropy}).  The \astropypkg
provides a full log of changes\footnote{\url{https://github.com/astropy/astropy/blob/stable/CHANGES.rst}}
over the course of the entire project and more details about individual
subpackages are available in the documentation.\footnote{\url{http://docs.astropy.org/en/stable/}}
Beyond what is mentioned below, most subpackages have seen increased performance
since the release of the v0.2 package.

% \subsection{Analytic Functions}
% moved to models and depreciated --
% just here for completeness at the moment
%\inlinecomment{BMS}{I suggest to order the subsection lexicographically as reading the draft the current order seemed very random (unlike at the docs page where they the grouping logic is obvious)}
%\inlinecomment{Tom R}{the only issue is if we rely on having e.g. units described first if they are used later}

@@

%\inlinecomment{Perry G}{Agreeing with Tom below, I would avoid using terms like subclass, decorators, mixin, namespace, (and perhaps even class, instead using objects consistently; this is no place to be pendantic about the distinction between classes and objects; even to the extent of adding a qualifying sentence somewhere like: "In this paper we will use object when we mean object or classes; in software there are important distinctions between the two terms, but they are not relevant in this paper."), and the like and replace them with more generic descriptions, e.g., "extended", or "tools to make defining functions that can use quantities simple to write" }
%\inlinecomment{Tom R}{I think saying 'subclass' here is too detailed and is mentioned below in 3.1.1 if needed - 'make it possible to attach units to numpy arrays' or something like that?}

As of \astropypkg version 2.0, units and quantities, prevalent in most of its
subpackages, have become a key concept for using the package as a whole.
Units are intimately entwined in the definition of astronomical coordinates;
Thus, nearly all functionality in the \texttt{astropy.coordinates} subpackage
(see \sectionname~\ref{sec:coordinates}) depends on them.
For most other subpackages, quantities are at least accepted and often expected
by default.

@@

%\subsubsection{Speed improvements}

%\inlinecomment{Tom R}{I wonder if we might just want to mention at the top level that performance has been improved in places? This seems a bit too detailed}

%The \texttt{Quantity} object is now significantly
        %faster than in early versions of \astropypkg. This is in large part due
        %to the fact that \texttt{Quantity} subclasses the \texttt{numpy.ndarray}
        %object, but even relative to more recent versions of \astropypkg,
        %operations with \texttt{Quantity} objects are generally faster. This is
        %largely thanks to improvements in the way \texttt{numpy} arrays can be
        %subclassed.

@@

\subsubsection{Defining functions that require quantities}
        When writing code or
        functions that expect \texttt{Quantity} objects, we often want to
        enforce that the input units have the correct physical type.
        For example, we may want to require only length-type \texttt{Quantity}
        objects.
        \texttt{astropy.units} provides a tool called \texttt{quantity\_input()}
        that can perform this varification automatically to avoid repetitive
        code.

@@

The classes that handle coordinate representations (the \texttt{Representation}
classes) act like three-dimensional vectors and thus support vector arithmetic.
The classes that represent reference systems and frames (the \texttt{Frame}
classes) internally use \texttt{Representation} objects to store the coordinate
data---that is, the \texttt{Frame} classes accept coordinate data, either as a
specified \texttt{Representation} object, or using short-hand keyword arguments
to specify the components of the coordinates.
These preferred representation and short-hand component names differ between
various astronomical reference systems.
For example, in the ICRS frame, longitude and latitude are right ascension
(\texttt{ra}) and declination (\texttt{dec}), whereas in the Galactic frame, the
spherical angles are Galactic longitude (\texttt{l}) and latitude (\texttt{b}).
Each \texttt{Frame} class defines its own component names and preferred \texttt{Representation} class.
The frame-specific component names map to corresponding components on the
underlying \texttt{Representation} object that internally stores the coordinate
data.
For most frames, the preferred representation is spherical, although this is
determined primarily by the common use in the astronomical community.

@@

\begin{figure}
\includegraphics[width=\textwidth]{coordinates_graph.pdf}
\caption{%
    The full graph of possible reference frame transformations implemented in
    \texttt{astropy.coordinates}.
    \label{fig:frame-transform-graph}
}
\end{figure}

@@

\subsubsection{Proper motion and velocity transformations}
    In addition to positional coordinate data, the \texttt{Frame} classes now
    also support velocity data.
    As the default representation for most frames is spherical, most of the
    \texttt{Frame} classes expect proper motion and radial velocity components
    to specify the velocity information.
    The names of the proper motion components all start with \texttt{pm} and
    adopt the same longitude and latitude names as the positional components.
    Transforming coordinates with velocity data is also supported, but in some
    cases the transformed velocity components have limited accuracy because the
    transformations are done numerically instead of analytically.
    \figurename~\ref{fig:frame-transform-graph} highlights which
    velocity transformations can be done exactly and which transformations are
    done using a finite-difference scheme.
    The low-level interface for specifying and transforming velocity data (see
    the next point) is currently experimental.
    As such, in v2.0, only the \texttt{Frame} classes (and not the
    \texttt{SkyCoord} class) support handling velocities.

%\subsubsection{Derivatives of coordinate representations}

%\inlinecomment{Tom R}{does this really warrant a separate section or is it just a byproduct of the velocity stuff that can be mentioned in 3.3.2?}

    %As mentioned above, the \texttt{Representation} classes act like
    %three-dimensional vectors.
    %\texttt{astropy.coordinates} now also supports handling first derivatives of
    %vectors / representations through the new \texttt{Differential} classes.
    %The \texttt{Differential} classes are currently used internally within the
    %\texttt{Frame} classes to store the velocity data.

\subsubsection{Solar System Ephemerides}
    Also new is support for computing ephemerides of major solar system bodies
    and outputting the resulting positions as coordinate objects.
    These ephemerides can be computed either using analytic approximations from
    ERFA or from downloaded JPL ephemerides (the latter requires the
    \package{jplephem}\footnote{\url{https://github.com/brandon-rhodes/python-jplephem}}
    optional dependency and an internet connection).
    % hamogu: I think the specific names of functions can be in the docs
    % and don't need to be repeated here.
    %The \texttt{get\_body} and \texttt{get\_body\_barycentric} functions provide
    %the gateway to this functionality, yielding \texttt{SkyCoord} objects given
    %a specific body and choice of ephemeris source.

@@

\tablename~\ref{tab:coordinate_benchmarks} gives an example of the relative
accuracy of the conversion from FK4 to Galactic coordinates for all pairs of
tools. This shows, for example, that \astropy, Kapteyn, and PyTPM agree
perfectly (to the precision shown), while PALpy, pySLALIB, and PyAST also
agree perfectly amongst themselves. However, there is an offset of around
0.2\arcsec\space with the former three packages. Finally, PyEphem disagrees
with other packages by 0.4--0.8\arcsec. These values are only meant to be
illustrative and will change over time as the benchmarks are refined and the
packages updated.
%\inlinecomment{Perry G}{Wouldn't the table be better represented as a matrices? Or perhaps as a matrix as a crude image with colors or intensities representing the size of the quantity. Is the difference between median and mean significant enough to warrant showing both? A link to the actual numbers may satisfy readers that want the exact values, but I suspect the majority don't need to see that precision.}
\begin{table}
\begin{center}
\begin{tabular}{llcccc}
\hline
\hline
Package 1 & Package 2 & Median & Mean & Maximum & Std. Dev. \\
          &           & arcsec & arcsec & arcsec & arcsec \\
\hline
astropy  &  kapteyn  & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata  &  palpy    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyast    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyephem  & 0.459  & 0.458 & 0.860 & 0.231 \\
\nodata  &  pyslalib & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pytpm    & 0.000  & 0.000 & 0.000 & 0.000 \\
kapteyn  &  palpy    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyast    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyephem  & 0.459  & 0.458 & 0.860 & 0.231 \\
\nodata  &  pyslalib & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pytpm    & 0.000  & 0.000 & 0.000 & 0.000 \\
palpy    &  pyast    & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata    &  pyephem  & 0.563  & 0.570 & 1.012 & 0.253 \\
\nodata    &  pyslalib & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata    &  pytpm    & 0.218  & 0.196 & 0.249 & 0.056 \\
pyast    &  pyephem  & 0.563  & 0.570 & 1.012 & 0.253 \\
\nodata    &  pyslalib & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata    &  pytpm    & 0.218  & 0.196 & 0.249 & 0.056 \\
pyephem  &  pyslalib & 0.563  & 0.570 & 1.012 & 0.253 \\
\nodata  &  pytpm    & 0.459  & 0.458 & 0.860 & 0.231 \\
pyslalib &  pytpm    & 0.218  & 0.196 & 0.249 & 0.056 \\
\hline
\end{tabular}
\end{center}
\caption{Comparison of the accuracy of the FK4 to Galactic transformation between different packages.\label{tab:coordinate_benchmarks}}
\end{table}

@@

The \texttt{NDDataBase} class provides the abstract interface for gridded data
with attributes for accessing metadata, the world-coordinate system (WCS),
uncertainty arrays matched to the shape of the data, and other traits.
Building on this interface, the \texttt{NDData} class provides a minimal
working implementation for storing \package{numpy} arrays. These classes serve
as useful base classes for package authors wishing to develop their own classes
for specific use cases and as containers for exchanging gridded data.

@@

The \package{astropy.table} subpackage provides functionality for representing
and manipulating heterogeneous data.
%\inlinecomment{APW:}{Need more introduction to the table subpackage!}
The package was described in detail in \cite{astropy}.
Thus, in what follows, we only summarize key new features or updates to
\package{astropy.table} since the previous \astropy paper.

%\inlinecomment{Perry G}{Some mention that some of these are inspired by Pandas?}

@@

A table can contain data that naturally form groups; For example, it may
contain multiple observations of a few sources at different points in time
and in different bands. Then, we may want to split the table into groups based
on the combination of source observed and the band, after which we combine the
results for each combination of source and band in some way (e.g., finding
the mean or standard deviation of the fluxes or magnitudes over time) or filter
the groups based on user-defined criteria. These kinds of grouping and
aggregation operations are now fully supported by \texttt{Table} objects.

@@

\subsubsection{Using astropy objects in tables}

The \texttt{Table} object now allows \texttt{Quantity}, celestial
coordinate (\texttt{SkyCoord}), and date/time (\texttt{Time}) objects to
be used as columns. It also provides a general way for other user-defined
objects to be used as columns. This makes it possible, for instance, to easily
represent catalogs of sources or time series as in \astropy, while having both the
benefits of the \texttt{Table} object (e.g., accessing specific rows/columns
or groups of them and combining tables) and of, for example,
the \texttt{SkyCoord} or the \texttt{Time} classes (e.g., converting the
coordinates to a different frame or accessing the date/time in the desired time scale).

@@

The \package{astropy.io} subpackage provides support for reading and writing
data to a variety of ASCII and binary file formats, such as a wide range of
ASCII data table formats, FITS, and VOTable.
It also provides a unified interface for reading and writing data with these
different formats using the \package{astropy.table} subpackage.
For many common cases this simplifies the process of file input and output and
reduces the need to master the separate details of all the I/O packages within
\astropypkg.

%\inlinecomment{SCO}{Table and FITS tables}

%\subsubsection{Unified file read/write interface}

%\inlinecomment{Check when it was added (0.3?). Overview of the supported formats (ASCII,FITS, votable, HTML, JSViewer/Datatables).}

% We summarize key new features or updates to \package{astropy.io} below,
% organized by sub-subpackage.

% \inlinecomment{Tom R}{is a sub-subpackage a thing, or is it just 'subpackage' all the way down?}

\subsubsection{ASCII}
% \subsubsection{Enhanced CSV format}

@@

% \subsubsection{\texttt{.ascii}: fast readers and writers}

The \package{astropy.io.ascii} subpackage now includes a significantly faster
Cython/C engine for reading and writing ASCII files. This is available for the
most common formats.  On average the new engine is about 4 to 5 times faster
than the corresponding pure-\python implementation, and is often comparable to
the speed of the \package{Pandas} \citep{pandas} ASCII file interface.  The
fast reader has parallel processing option that allows harnessing multiple
cores for input parsing to achieve even greater speed gains.  By default,
\texttt{read()} and \texttt{write()} will attempt to use the fast C engine
when dealing with compatible formats. Certain features of the full read
/ write interface are not available in the fast version, in which case the
pure-\python version will automatically be used.

% \subsubsection{\texttt{.ascii}: HTML tables}

The \package{astropy.io.ascii} subpackage now provides the capability
to read a table within an HTML file or web URL into an astropy
\texttt{Table} object. Conversely a \texttt{Table} object can now
be written out as an HTML table.

@@

\package{matplotlib} allows a custom normalization and stretch to be used when displaying data by passing a normalization object. The \package{astropy.visualization} package also provides a normalization class that wraps the interval and stretching objects into a normalization object that \package{matplotlib} understands.
